---
title: "ps1"
format: pdf
editor: visual
---

## 2. Get to Know Your Data {style="line-height: 1.2;"}

1.  **Load the Tidyverse and read in your acs data.**

```{r}
library(tidyverse)
acs <- read_csv("data/raw/acs2024.csv")
```

2.  **Using the IPUMS documentation (online at https://usa.ipums.org), look up what each variable in your dataset means. Pay special attention to TRANTIME (What does a value of 0 mean?) and PERWT (What does this represent?).**

According to the IPUMS website, the 2024 ACS 1-year sample is the American Community Survey from 2024. The variables chosen from this dataset are as described below (the variables with missing information lack readily avaialble explanations on the IPUMS site and, thus, will not be used in analysis):

-   YEAR represents the date each respondent filled out the survey. Since this data is from 2024, all respondents have 2024 as their YEAR variable.

-   SAMPLE tells us which sample the respondent's data was pulled from.

-   SERIAL indicates which household the respondent belongs to.

-   CBSERIAL indicates the unique household identification number, as assigned by the Census Bureau. All persons in a household are assigned the same number.

-   HHWT tells us how many households in the US are represented by a given household in the sample.

-   CLUSTER

-   STATEFIP tells us which state the respondent's household was located, using the Federal Information Processing Standards coding scheme.

-   GQ

-   PERNUM numbers all persons within each household consecutively in the order in which they appear on the original census or survey form.

-   PERWT tells us how many people in the US are represented by a given respondent in the sample.

-   SEX reports the respondent's assigned sex at birth: (1) male or (2) female. (9) is reserved for respondent's with missing/unreported data.

-   AGE reports the respondent's age in years.

-   EDUC tells us the level of education attained by the respondent. Each level is coded numerically as follows: (0) NA or no schooling, (1) Nursery school up to grade 4, (2) Grades 5, 6, 7, or 8, (3) Grade 9, (4) Grade 10, (5) Grade 11, (6) Grade 12, (7) 1 year of college, (8) 2 years of college, (9) 3 years of college, (10), 4 years of college, (11) 5+ years of college, and (99) Missing.

-   EDUCD

-   EMPSTAT reports whether the respondent was a part of the labor force: (0) NA, (1) Employed, (2) Unemployed, (3) Not in the labor force, and (9) Missing.

-   EMPSTATD

-   TRANTIME tells us the total amount of time, in minutes, that it takes the respondent to travel from home to work on average in the past week. A value of 0 means that the data is missing from the data set.

-   INCTOT reports the respondent's total pre-tax income or losses from all sources for the previous year.

3.  **Create new variables from the raw IPUMS codes.**

```{r}
acs_clean1 <- acs |>
  mutate(female = if_else(SEX == 2, 1, 0)) |>
  mutate(less_than_hs = if_else(EDUC <= 2, 1, 0),
         hs_only = if_else(EDUC %in% c(3, 6), 1, 0),
         some_college = if_else(EDUC %in% c(7, 9), 1, 0),
         college_only = if_else(EDUC == 10, 1, 0),
         advanced_degree = if_else(EDUC == 11, 1, 0)) |>
  mutate(employed = if_else(EMPSTAT == 1, 1, 0),
         unemployed = if_else(EMPSTAT == 2, 1, 0),
         not_in_labor_force = if_else(EMPSTAT == 3, 1, 0))
```

4.  **Why does this matter? In 2–3 sentences, explain what could go wrong if you accidentally treated “N/A” codes as real zeros when calculating summary statistics. Give a concrete example using one of the variables in your dataset.**

If you accidentally treated "N/A" codes as real zeros when calculating summary statistics then those missing responses would be incorrectly interpreted as meaningful values. For example, since TRANTIME uses 000 to indicate missing data, those zeros should not be used to calculate the average commute time. If they included, they would lower the mean commute time, making it seem like people have shorter average commutes than they actually do.

5.  **Create a summary statistics table for your dataset.**

```{r}
library(knitr)

acs_clean2 <- acs_clean1 |>
  mutate(commute_time = if_else(TRANTIME %in% c(000, 888), NA_real_, TRANTIME),
         income = if_else(INCTOT %in% c(9999998, 9999999), NA_real_, INCTOT))

sumstats <- function(x) {tibble(N = sum(!is.na(x)),
                                Mean = mean(x, na.rm = TRUE),
                                Median = median(x, na.rm = TRUE),
                                "Std. Dev." = sd(x, na.rm = TRUE),
                                Min = min(x, na.rm = TRUE),
                                Max = max(x, na.rm = TRUE))}

table1 <- bind_rows(sumstats(acs_clean2$AGE) |>
                      mutate(Variable = "Age"),
                    sumstats(acs_clean2$female) |>
                      mutate(Variable = "Female"),
                    sumstats(acs_clean2$less_than_hs) |>
                      mutate(Variable = "Less than High School"),
                    sumstats(acs_clean2$hs_only) |>
                      mutate(Variable = "High School Only"),
                    sumstats(acs_clean2$some_college) |>
                      mutate(Variable = "Some College"),
                    sumstats(acs_clean2$college_only) |>
                      mutate(Variable = "College Only"),
                    sumstats(acs_clean2$advanced_degree) |>
                      mutate(Variable = "Advanced Degree"),
                    sumstats(acs_clean2$employed) |>
                      mutate(Variable = "Employed"),
                    sumstats(acs_clean2$unemployed) |>
                      mutate(Variable = "Unemployed"),
                    sumstats(acs_clean2$not_in_labor_force) |>
                      mutate(Variable = "Not in Labor Force"),
                    sumstats(acs_clean2$commute_time) |>
                      mutate(Variable = "Commute Time (mins)"),
                    sumstats(acs_clean2$income) |>
                      mutate(Variable = "Total Income ($)")) |>
  select(Variable, N, Mean, Median, "Std. Dev.", Min, Max)

kable(table1,
      caption = "Table 1: Summary Statistics for 2024 ACS Sample",
      digits = 2,
      booktabs = TRUE)
```

6.  **Does the maximum commute time make sense for a daily commute?**

Though long, the maximum commute time make sense. Its over three hours, which is quite a distance, but especially for people commuting into metropolitan areas, a commute time of this length is definitely not implausible.

## 3. Who Should Be in Your Analysis {style="line-height: 1.2;"}

1.  Investigate the Data

```{r}
acs_clean3 <- acs_clean2 |>
  summarize(total = n(),
            no_commute = sum(TRANTIME == 0, na.rm = TRUE),
            percent_zero = 100 * mean(TRANTIME == 0, na.rm = TRUE))
```

A. A transit time of 000 means that the respondent's data is missing or not relevant. 2,062,945 (60.3%) of the 3,422,888 total respondents had a transit time of 000.

```{r}
acs_clean2 |>
  filter(TRANTIME == 000) |>
  count(EMPSTAT)
```

B. Most of the respondents without commute times are not in the labor force (3). Of those employed, 248,078 do not commute, and we can, thus, assume that they work from home.

2.  Make your decision.

```{r}
commuters <- acs_clean2 |>
  filter(EMPSTAT == 1,
         commute_time > 0,
         AGE >= 16)
```

My criteria for the commuters variable is to limit the variable to only those that are employed and have a commute time greater than zero (to weed out remote workers). Those with commute times also should inherently be those outside of school, but just to make sure that children were not included in this sample, I filtered for only those above the age of 16.

3.  Verify your subset.

```{r}
nrow(commuters)
min(commuters$commute_time, na.rm = TRUE)
```

To verify my subset, I tried to find the number of rows (the volume of individual participants) and the minimum transit time. The number of commuters in our dataset is 1,359,943 and their minimum transit time is one minute.

4.  Create a new summary statistics table for your commuters subset, using the same format as your table in Section 2.

```{r}
table3_4 <- bind_rows(sumstats(commuters$AGE) |>
                      mutate(Variable = "Age"),
                    sumstats(commuters$female) |>
                      mutate(Variable = "Female"),
                    sumstats(commuters$less_than_hs) |>
                      mutate(Variable = "Less than High School"),
                    sumstats(commuters$hs_only) |>
                      mutate(Variable = "High School Only"),
                    sumstats(commuters$some_college) |>
                      mutate(Variable = "Some College"),
                    sumstats(commuters$college_only) |>
                      mutate(Variable = "College Only"),
                    sumstats(commuters$advanced_degree) |>
                      mutate(Variable = "Advanced Degree"),
                    sumstats(commuters$employed) |>
                      mutate(Variable = "Employed"),
                    sumstats(commuters$unemployed) |>
                      mutate(Variable = "Unemployed"),
                    sumstats(commuters$not_in_labor_force) |>
                      mutate(Variable = "Not in Labor Force"),
                    sumstats(commuters$commute_time) |>
                      mutate(Variable = "Commute Time (mins)"),
                    sumstats(commuters$income) |>
                      mutate(Variable = "Total Income ($)")) |>
  select(Variable, N, Mean, Median, "Std. Dev.", Min, Max)

kable(table3_4,
      caption = "Table 2: Summary Statistics for Commuters in the 2024 ACS Sample",
      digits = 2,
      booktabs = TRUE)
```

## 4. Visualize and Interpret {style="line-height: 1.2;"}

1.  Create a histogram of commuters using ggplot2.

```{r}
library(ggplot2)

commute_histogram <- ggplot(commuters, aes(x = TRANTIME)) +
         geom_histogram(binwidth = 5, 
                        color = "white", 
                        fill = "pink") +
         labs(title = "Commute Time Distribution (Min)",
              x = "Commute Time (Min)",
              y = "Number of Commuters") +
  theme_minimal()
```

2.  In 2–3 sentences, describe what you see: Is it symmetric or skewed? Where is the center?
    Are there any notable features (e.g., spikes at certain values)?

According to the graph, most commuters are clustered between 10 and 30, with the center seemingly around 20 minutes. There is a noticeable spike around 15 and 30 minutes. Overall, the data is definitely not symmetrical and has a long tail ending around 155 minutes at the maximum commute time.

3.  Compare the mean and median commute time.

From the table, the mean commute time (27.22 minutes) is larger than the median commute time (20 minutes). This tells us that the data is right-skewed, which means that a small number of longer commute times pull the average higher than the center of the data. The numbers here are consistent with the graph.

4.  Save your histogram.

```{r}
ggsave(filename = "output/commute_histogram.png", 
       width = 8,
       height = 5)
```
